---
title: "Final Data Documentation"
author: "Cindy Wang"
date: "2023-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
Data_Science_Salary_Data <- read.csv("Final.csv")
head(Data_Science_Salary_Data)
```

## Original Data Sources:

1. Data Jobs Salaries from Kaggle, and the data is collected from the website https://ai-jobs.net/salaries/download/, its data gathered from professionals and employers all over the world. 

2. Employment Rate from OECD website

## Variables

#### year: 

The year the salary was paid               

#### experience_level:

The experience level in the job during the year with the following possible values:

EN

Entry-level / Junior

MI

Mid-level / Intermediate

SE

Senior-level / Expert

EX

Executive-level / Director

#### employment_type: 

The type of employement for the role:

PT

Part-time

FT

Full-time

CT

Contract

FL

Freelance

#### job_title:

The role worked in during the year

#### salary:

The total gross salary amount paid

#### salary_currency:

The currency of the salary paid as an ISO 4217 currency code

#### salary_in_usd:

The salary in USD

#### employee_residence:

Employee's primary country of residence in during the work year as an ISO 3166 country code

#### remote_ratio:

The overall amount of work done remotely, possible values are as follows:

0

No remote work (less than 20%)

50

Partially remote/hybird

100

Fully remote (more than 80%)

#### company_location:

The country of the employer's main office or contracting branch as an ISO 3166 country code

#### company_size:

The average number of people that worked for the company during the year:
S
less than 50 employees (small)
M
50 to 250 employees (medium)
L
more than 250 employees (large)

#### employment_rate:

Measure of the extent to which available labor resources are being used

#### salary_range:

Salary in USD rounded

#### remotegroup:

Remote working for each company

```{r, eval=FALSE, echo=FALSE}
nrow(Data_Science_Salary_Data)
ncol(Data_Science_Salary_Data)
sum(is.na(Data_Science_Salary_Data))
```

## Summary Statistics:

The data set has 481 rows, 15 columns, and there are no missing values.




